Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2022-01-26 fstats_pca_lewontin.tex
proj figure

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_admixture_pca.pdf}
	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot. A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ (with proportion $1-\alpha$)and $X_3$ (proportion ($\alpha$), with $X_2$ contributing proportion $\alpha$. Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$. B: PCA representation of the scenario in A. $X_y$ originates on the segment connecting $X_2$ and $X_3$, and subsequent drift may move it in a random direction. C: $F_3(X_x; X_2, X_3)$ and negative region (light gray circle). $F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D). }
	\label{fig:admix}
\end{figure}

@@

@@

\paragraph{$F_3$ on a 2D PCA-plot.} 
 If we project this $n$-ball on a two-dimensional plot, $\overline{X_2X_3}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot. This geometry is perhaps easiest visualized on a globe. If we look at the globe from a view point parallel to the equator, both the north and south poles are visible at the very edge of the circle. But if we look at it from above the north pole, the north- and south-poles will be at the very same point.

 Thus if $\hat{F}_3 \ll F_3$, the ``true'' circle will be bigger than what would be predicted from a 2D-plot, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics. This is because they are outside the $n$-ball in higher dimensions. The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix).

>>>>

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_admixture_pca.pdf}
	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot. A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ (with proportion $1-\alpha$)and $X_3$ (proportion ($\alpha$), with $X_2$ contributing proportion $\alpha$. Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$. B: PCA representation of the scenario in A. $X_y$ originates on the segment connecting $X_2$ and $X_3$, and subsequent drift may move it in a random direction. C: Negative region (light gray) for $F_3(X_x; X_2, X_3)$ and for $\hat{F}_3^{(2)}(X_x; X_2, X_3)$ based on two dimensions (dark gray). D: $F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D). }
	\label{fig:admix}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.4\textwidth]{figures/fig_proj.pdf}
	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: $F_3(X_x; X_1, X_2)$ in two dimensions (full data). Points like $X_x$ inside the circle (red) have $F_3 <0$, but points outside ($X_3, X_4$, blue) have positive $F_3$-statistics. B: Projection on one dimension. The full rejection region (light gray bar) is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar). Points with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray). $X_3$ also projects onto the gray bar, even though its $F_3$ is positive. Points such as $X_4$ that project outside the rejection region have $F_3>0$.}
	\label{fig:proj}
\end{figure}

@@

\paragraph{$F_3$ and dimensionality reduction} 
 If we project this $n$-ball on a two-dimensional plot, $\overline{X_2X_3}$ will usually not align with the PCs; thus the ball may be somewhat larger than it appears on the plot. This geometry is perhaps easiest visualized on the example of projecting $F_3(X_x; X_1, X_2)$ from a two-dimensional space onto a single dimension (Figure \ref{fig:proj}). In that example, the distance between $X_1$ and $X_2$ has a substantial contribution from PC2, and so the negative region (light gray) is larger than what would be predicted from just one dimension (dark gray bar), but if $\hat{F}_3 \approx F_3$, the two areas would be very close. Thus, if considering a reduced-dimension PCA-plot, some points (such as $X_3$) may project inside the negative region, but have positive $F_3$ because they are outside the $n$-ball in higher dimensions. The converse interpretation is more strict: if a population lies outside the circle on \emph{any} projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix). An intuitive example is given by $X_4$ in Figure \ref{fig:proj}: all points projecting to the same point on Figure \ref{fig:proj}B as $X_4$ lie outside the circle.
